Tools for Identifying Partisan Gerrymandering (PA)

### 2019-06-18 Latex/_maintext.tex
Update _maintext.tex

    \lettrine[lines=3, lraise=0.1, nindent=0em]{\color{lightgray}I}{n} \textit{League of Women Voters v. Pennsylvania}, decided January 22, 2018 (henceforth abbreviated \textit{LWV}), the Pennsylvania Supreme Court offered a new and distinctive test for unconstitutional partisan gerrymandering under which it found the state’s congressional districts to be a ``severe and durable" partisan gerrymander that violated a long-standing provision of the state’s constitution. The congressional map that was struck down was drawn in 2011 by a Republican-controlled legislature and signed into law by the then Republican governor. It was used in elections from 2012 to 2016. Prior to the \textit{LWV} decision, many academics had identified this map as a partisan gerrymander \citep[see e.g.,][]{Royden2017}, with some even identifying it as the most egregious congressional districting in the U.S. \citep[][ esp. Figure 3]{Wang2016_SLR}.

@@

    The Pennsylvania Supreme Court held that ``the fact that the 2011 Plan cannot, as a statistical matter, be a plan directed at complying with traditional redistricting requirements is \underline{sufficient} to establish that it violates the Free and Equal Elections Clause" [of the Pennsylvania State Constitution] (slip op. at p.128, emphasis added).

@@

        \footnote{``Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage." Pa. Const. art. I, § 5.}

@@

    However, the Court also considered a variety of other indicators involving the partisan aspects of the plan in coming to the conclusion that the map created a ``severe and durable" partisan gerrymander.

@@

        \footnote{The trial court considered a variety of factors offered in expert witness testimony, including evidence about the map‘s political impact. The trial court’s findings of fact reveal that the 2011 map exhibited clearly a multiplicity of indicia of partisan gerrymandering. In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander. The trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered. However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents. Instead, it expressed a pious hope that the U.S. Supreme Court, in its review of the appeals of lower court cases, such as the 2017 case of \href{https://www.brennancenter.org/sites/default/files/legal-work/16-1161_Opinion.pdf}{\textit{Gill v. Whitford 585 U. S. \_\_\_ (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply. The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}

@@

        \footnote{On March 20, 2018, Rep. Cris Dush, of Jefferson County introduced a measure that would impeach the four Democratic members of the court that voted to replace the map. His argument is that the members of the court overstepped their rights and violated the Separation of Powers principle of the state. Others saw this as simply sour grapes: \\ \href{https://wapo.st/2LlEVvE}{``Pennsylvania Republicans lost the redistricting battle. Now, they’re declaring war on the courts", https://wapo.st/2LlEVvE (Christopher Ingraham, February 22, 2018).;} \\ \href{https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html}{``Pa. Republicans are talking about impeaching state Supreme Court justices. Do they have an argument?", https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html (Andrew Seidman \& Jonathan Lai, February 22, 2018).}}

@@

    In that remand decision (\textit{Gill v. Whitford, 585 U.S. \_\_\_ (2018)}), the Supreme Court indicated that any standard for partisan gerrymandering must be a challenge to individual districts which demonstrates that the voters in the particular district have had their votes either ``packed" or ``cracked."

@@

    Districts that have been \textit{packed} are those which have had voters from one partisan leaning grouped together so as to waste votes of the group because they win the district by an overwhelming majority, thus preventing those excess votes from being used more efficiently in other districts. Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district. The standard for unconstitutional gerrymandering at the federal level -- if there is ever is one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the U.S. Supreme Court appears to be insisting on a district-specific test of packing and cracking. 

@@

    Some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts. Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them. Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard. Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked" to address what \citet[][see also \citep{Grofman_Cervas_2018_ELJ}]{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact. The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map. 

@@

        \footnote{See NYT Upshot: \href{https://nyti.ms/2w8b19k}{``Democrats Didn’t Even Dream of This Pennsylvania Map. How Did It Happen?", https://nyti.ms/2w8b19k (Nate Cohn, Feb. 21, 2018).;} \\
        \href{https://nyti.ms/2wkpEFT}{``Hundreds of Simulated Maps Show How Well Democrats Fared in Pennsylvania", https://nyti.ms/2wkpEFT (Nate Cohn, Feb. 26, 2018).}}

@@

    All the plans create contiguous districts. Similarly, all four plans appear to satisfy voting rights requirements. Thus, we do not discuss these two ``good government'' criteria further. Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole. 

@@

        \subsection*{Measurement and Data}

@@

    \input{Figures/_maps.tex}

@@

    The measurement of compactness has long been a central issue in the districting literature \citep[see e.g.,][]{Reock1961, Niemi1990, Polsby1991, Webster2013, Kaufman2018}. Indeed, the original gerrymander was assumed to be evidence of political manipulation because of its irregular shape \citep{Martis2008}. Some minimal level of compactness can make it harder to gerrymander \citep[][cf. \citealt{Webster2013}]{Reock1961, Altman1998}. But compactness is not a magic wand that rules out the possibility of political manipulation. Moreover, as many authors have shown, compactness has multiple dimensions, and these different dimensions need not move in parallel with one another; thus a plan appears compact on one dimension might appear ill-compact on another. The two most common ideas of compactness refer, on the one hand, to how close a legislative district’s boundaries are to its geographic center and, on the other, to how ``regular" in shape a district appears to be \citep{Niemi1990, Kaufman2018}. In Table \ref{tab:summaries} we report two well-known measures which tap respectively each of these two dimensions. The \textit{Polsby-Popper} measure looks at perimeter irregularity by examining the area of the district compared to that of a circle with similar perimeter, while the \textit{Reock} measure compares the area of a district with that of the district’s circumscribing circle \citep{Reock1961, Polsby1991}. 

@@

    The 2011 Enacted map looks awful by both compactness measures compared to the Court-ordered map, or even compared to the two proposed remedial maps; while both the Republican and the Democratic proposed remedial maps look reasonable by both compactness measures, even though the Democratic plan is superior to the Republican plan and the court-ordered map is superior to both. However, until we examine the likely partisan effects of these four plans we are not in a position to conclusively rule out the possibility that one or more of the remedial plans is a ``\textit{stealth gerrymander}". As is now well recognized, gerrymandering can be found even in plans with compact appearing districts, or in a plan with districts that preserve most county borders.

@@

    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map. And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map. Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally'' tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters (see below). Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised'' pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander'' -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".

@@

    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, Figure \ref{fig:maps} shows all four maps in the same scale, in a gray and white format. What is visually apparent from these maps is how aesthetically ``ugly" the 2011 map is compared to all of its alternatives. But it also clear that three remedy maps that are similar in having relatively few county cuts nonetheless can look quite different in terms of how each configures districts.

@@

@@

    Now we turn to a consideration of probable partisan consequences of the four maps. We begin with a discussion of the 2011 map.
            \subsubsection*{Partisan results in actual elections held under the 2011 map}

@@

    Thus, the discrepancy between Republican vote share and Republican seat share was huge, and yet the seat share appeared to be immutable (see Table \ref{tab:congsum}).

@@

        \subsection*{Asymmetry measures for partisan outcomes in the 2011 Plan based on statewide election outcomes}
    Table \ref{tab:congsum} shows other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below). Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016. 

@@

    \underline{Positive} values favor Republicans , while \underline{negative} values favor Democrats.   

@@

    In our view, none of these four measures standing alone can be taken to be proof of gerrymandering. Rather we see them as potentially reinforcing indicators. Our own analyses have found hypotheticals where these measures can generate either a false positive or a false negative, but we have also found that consistent bias in the same direction is unlikely when we use multiple indicators (data omitted for space reasons). Here, however, all four indicators, like the actual election results, point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan -- namely that it is a blatant pro-Republican partisan gerrymander.
        \subsection*{Evaluating Partisan Gerrymandering using Statewide elections to provide a normal vote baseline}
        \footnote{We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level (see discussion on alternative ways to tabulate vote-shares in \citet{Grofman_et_al_1997__IntedgratedPerspective_ES}. Both ways result in the same district partisanship outcomes in all plans.}
    We use these five elections to calculate the same four previously introduced metrics but now for all four plans.
    \footnote{We report the Republican share of the two-party vote ($\delta_{i}$). Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $. To find the overall state-wide vote share, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $. By averaging, we reduce the influence of turnout variation among districts. This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf. \cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}. In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes. In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades \citep[cf.][]{Jacobson2015}.}
    We aggregate voting district level (precincts) data for these five statewide elections to the congressional district level.
       \footnote{We wrote a function in \texttt{R} to assign all VTDs to a district based on the centroid of the VTD. This method creates a slight difference between it and the actual vote, but averages just -0.06\% per district. Most of the mis-allocation is between two adjacent districts, District 1 and District 13, both overwhelmingly Democratic districts.}
    We do need to check that our five-election composite offers a plausible basis for a prediction model once we take into account electoral tides. 
%       \footnote{We can only validate our measure for the 2016 election and only for the 2011 enacted plan. For all the remedial plans we are analyzing, the composite is our ``best guess". We have to make some assumptions about state and district inter-election swings. For simplicity, we assume that all districts have a national shift equal to the state-level swing, ie, difference from the 2016 elections. Later in the paper, we conduct simulated elections that vary the district level variability.}
    Encouragingly, our five-election composite results is the same 13R-5D result that the 2016 congressional races delivered.  
%       \footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}
    Additionally, each district's outcome is identical to the 2016 congressional district results, when we apply a uniform swing to match the results from the state-wide elections to the results from the congressional elections. 
    We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a ``normal vote" baseline that must be ``corrected" for incumbency effects if we wish to actually predict outcomes.
%        \footnote{After uniformly shifting each district's composite such that the statewide average equalled the actual 2018 statewide average. The large range is indicative of the number of competitive districts that Republican candidates won. Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won re-election by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely predicted outcome.}
    We must distinguish our ability to assess asymmetry in expected seats-vote relationships based on the ``normal vote" distribution of party support across districts -- and thus to assess the degree to which a plan is a gerrymander -- from our ability to predict actual election outcomes. The more competitive districts there are in a map, the more results depend strongly on electoral tides, and the greater the potential for winning vote margins in some districts to be small and determined by factors idiosyncratic to a district and by incumbency advantage. The 2011 map has 13 non-competitive districts (10+ percentage point margin) with a clear 9-4 Republican advantage, thus virtually guaranteeing continued Republican dominance regardless of aggregate two party vote share. The Court drawn map, comparatively, had only 8 non-competitive districts, with a balance of 3-R and 5-D. Which party will control the delegation under a highly competitive plan such as the Court-ordered plan depends almost entirely on the state-wide vote share and on incumbency advantage, along with idiosyncratic features of the competitive districts that might be small in magnitude but still matter in highly competitive seats. 
    In the congressional elections of 2018, the raw Republican two-party vote share in 2018 was 44.8\%, which represented a Democratic advantage, but not enough to overcome the twin barriers to a Democratic majority of over-concentration of Democratic support in urban areas and continued Republican incumbency advantage.
    When we look at the five election composite we need to (a) adjust it for the actual statewide congressional Democratic vote share in 2018, and (b) to take into account this differential incumbency advantage that still benefits Republicans. Without doing so, the outcomes of the Court-ordered plan have a wide range, with anywhere from a 5R to an 11R outcome possible because of the high number of competitive seats. However, when we take both factors into account by uniformly shifting the state-wide composite to reflect the 2018 statewide two-party congressional vote, and by adding in congressional incumbency advantage, which we have estimated in the 2012-2016 period to be on the order of magnitude of 4.3\% (analyses omitted for space reasons), we get a 9R-9D outcome as the most likely outcome in 2018 under the court-ordered map in our predictive model. Thus, our confidence in the five election projection method is further buttressed when we applied this method to predict the results of the 2018 election under the court-ordered remedial map in the light of the actual statewide aggregate results in 2018.

@@

    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats. Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%. At that point, both parties should, by symmetry, receive identical seats shares. Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias. Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero \underline{does not} imply proportional representational. What partisan bias taps is the differential treatment of parties. Thus, for example, if party R gets 61\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 61\% of the seats were it to have 52\% of the vote.

@@

        \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias (and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and reviewed in \citet{GrofmanKing2007_ELJ}). A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}   

@@

        \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature. In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections. \citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%. We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election. All three variations provide similar estimates, as do the results from \texttt{JudgeIt}. Readers should note that, with only 18 districts, too much weight should not be placed on the exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures. Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay. We have devoted more space to partisan bias than to the other measures of gerrymandering reported in this essay due to it's complexity than to the other more straightforward measures.} 

@@

    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}). Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans. We do this for each of the 1,000 simulated elections for each of the maps, and take the average.

@@

@@

            \subsubsection*{Measuring Gerrymandering}

@@

@@

    In the next section we introduce a potentially important distinction between ``\textit{neutral plans}" and ``\textit{fair plans}", and we consider some important aspects of Pennsylvania's electoral geography.

>>>>

    \lettrine[lines=3, lraise=0.1, nindent=0em]{\color{lightgray}I}{n} \textit{League of Women Voters v. Pennsylvania}, decided January 22, 2018 (henceforth abbreviated \textit{LWV}), the Pennsylvania Supreme Court offered a new and distinctive test for unconstitutional partisan gerrymandering under which it found the state’s congressional districts to be a ``severe and durable'' partisan gerrymander that violated a long-standing provision of the state’s constitution. The congressional map that was struck down was drawn in 2011 by a Republican-controlled legislature and signed into law by the then Republican governor. It was used in elections from 2012 to 2016. Prior to the \textit{LWV} decision, many academics had identified this map as a partisan gerrymander \citep[see e.g.,][]{Royden2017}, with some even identifying it as the most egregious congressional districting in the U.S. \citep[][ esp. Figure 3]{Wang2016_SLR}.

@@

    The Pennsylvania Supreme Court held that ``the fact that the 2011 Plan cannot, as a statistical matter, be a plan directed at complying with traditional redistricting requirements is \underline{sufficient} to establish that it violates the Free and Equal Elections Clause'' [of the Pennsylvania State Constitution] (slip op. at p.128, emphasis added).

@@

        \footnote{``Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage.'' Pa. Const. art. I, § 5.}

@@

    However, the Court also considered a variety of other indicators involving the partisan aspects of the plan in coming to the conclusion that the map created a ``severe and durable'' partisan gerrymander.

@@

        \footnote{The trial court considered a variety of factors offered in expert witness testimony, including evidence about the map‘s political impact. The trial court’s findings of fact reveal that the 2011 map exhibited clearly a multiplicity of indicia of partisan gerrymandering. In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander. The trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered. However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents. Instead, it expressed a pious hope that the U.S. Supreme Court, in its review of the appeals of lower court cases, such as the 2017 case of \href{https://www.brennancenter.org/sites/default/files/legal-work/16-1161_Opinion.pdf}{\textit{Gill v. Whitford 585 U. S. \underline{\hspace{3em}} (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply. The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}

@@

        \footnote{On March 20, 2018, Rep. Cris Dush, of Jefferson County introduced a measure that would impeach the four Democratic members of the court that voted to replace the map. His argument is that the members of the court overstepped their rights and violated the Separation of Powers principle of the state. Others saw this as simply sour grapes: \\ \href{https://wapo.st/2LlEVvE}{``Pennsylvania Republicans lost the redistricting battle. Now, they’re declaring war on the courts'', https://wapo.st/2LlEVvE (Christopher Ingraham, February 22, 2018).;} \\ \href{https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html}{``Pa. Republicans are talking about impeaching state Supreme Court justices. Do they have an argument?'', https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html (Andrew Seidman \& Jonathan Lai, February 22, 2018).}}

@@

    In that remand decision (\textit{Gill v. Whitford, 585 U.S. \underline{\hspace{3em}} (2018)}), the Supreme Court indicated that any standard for partisan gerrymandering must be a challenge to individual districts which demonstrates that the voters in the particular district have had their votes either ``packed'' or ``cracked.''

@@

    Districts that have been \textit{packed} are those which have had voters from one partisan leaning grouped together so as to waste votes of the group because they win the district by an overwhelming majority, thus preventing those excess votes from being used more efficiently in other districts. Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district. The standard for unconstitutional gerrymandering at the federal level -- if there is ever one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the U.S. Supreme Court appears to be insisting on a district-specific test of packing and cracking. 

@@

    Some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts. Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them. 
    Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard. Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked'' to address what \citet[][see also \citeyear{Grofman_Cervas_2018_ELJ}]{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact. The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map. 

@@

        \footnote{See NYT Upshot: \href{https://nyti.ms/2w8b19k}{``Democrats Didn’t Even Dream of This Pennsylvania Map. How Did It Happen?'', https://nyti.ms/2w8b19k (Nate Cohn, Feb. 21, 2018).;} \\
        \href{https://nyti.ms/2wkpEFT}{``Hundreds of Simulated Maps Show How Well Democrats Fared in Pennsylvania'', https://nyti.ms/2wkpEFT (Nate Cohn, Feb. 26, 2018).}}

@@

    Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole.
    All the plans create contiguous districts. Similarly, all four plans appear to satisfy voting rights requirements. Thus, we do not discuss these two ``good government'' criteria further.

@@

        \subsection*{Institutional Measures}

@@

    \input{Figures/_fig_maps.tex}

@@

    The measurement of compactness has long been a central issue in the districting literature \citep[see e.g.,][]{Reock1961, Niemi1990, Polsby1991, Webster2013, Kaufman2018}. Indeed, the original gerrymander was assumed to be evidence of political manipulation because of its irregular shape \citep{Martis2008}. Some minimal level of compactness can make it harder to gerrymander \citep[][cf. \citealt{Webster2013}]{Reock1961, Altman1998}. But compactness is not a magic wand that rules out the possibility of political manipulation. Moreover, as many authors have shown, compactness has multiple dimensions, and these different dimensions need not move in parallel with one another; thus a plan appears compact on one dimension might appear ill-compact on another. The two most common ideas of compactness refer, on the one hand, to how close a legislative district’s boundaries are to its geographic center and, on the other, to how ``regular'' in shape a district appears to be \citep{Niemi1990, Kaufman2018}. In Table \ref{tab:summaries} we report two well-known measures which tap respectively each of these two dimensions. The \textit{Polsby-Popper} measure looks at perimeter irregularity by examining the area of the district compared to that of a circle with similar perimeter, while the \textit{Reock} measure compares the area of a district with that of the district’s circumscribing circle \citep{Reock1961, Polsby1991}. 

@@

    The 2011 Enacted map looks awful by both compactness measures compared to the Court-ordered map, or even compared to the two proposed remedial maps; while both the Republican and the Democratic proposed remedial maps look reasonable by both compactness measures, even though the Democratic plan is superior to the Republican plan and the court-ordered map is superior to both. However, until we examine the likely partisan effects of these four plans we are not in a position to conclusively rule out the possibility that one or more of the remedial plans is a ``\textit{stealth gerrymander}''. As is now well recognized, gerrymandering can be found even in plans with compact appearing districts, or in a plan with districts that preserve most county borders.

@@

    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map. And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map. Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally'' tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters (see below). Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised'' pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander'' -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy''.

@@

    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, Figure \ref{fig:maps} shows all four maps in the same scale, in a gray and white format. What is visually apparent from these maps is how aesthetically ``ugly'' the 2011 map is compared to all of its alternatives. But it also clear that three remedy maps that are similar in having relatively few county cuts nonetheless can look quite different in terms of how each configures districts.

@@

            \subsection*{Asymmetry Measures}
    We emphasize that a partisan gerrymandering claim based on statewide political consequences should also provide evidence that the (a) gerrymandering indicators that are found cannot be explained simply by the geographic distribution of each party’s electoral support (e.g., differential concentration of electoral support can lead to what has been called a ``natural gerrymander''), and (b) that the partisan asymmetry in translating votes into seats is not due to chance alone (e.g., many highly competitive seats won by very narrow margins) but can be expected to be durable. With only 18 districts, sensitivity to chance effects is particularly important. 

@@

@@

@@

    There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below). Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016.

@@

    \underline{Positive} values favor Republicans, while \underline{negative} values favor Democrats. 

@@

    We now define these four measures of gerrymandering in which we will evaluate the four competing maps.

@@

    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats. Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%. At that point, both parties should, by symmetry, receive identical seats shares. Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias. Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero does \underline{not} imply proportional representational. What partisan bias taps is the differential treatment of parties. Thus, for example, if party R gets 61\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 61\% of the seats were it to have 52\% of the vote.

@@

        \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias (and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and reviewed in \citet{GrofmanKing2007_ELJ}). A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997}, and \citet{Johnston2002}.}   

@@

        \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature. In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections. \citet{Kastellec_et_al_2008_PS} similarly simulates data for average vote shares between 45\% and 55\%. We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election. All three variations provide similar estimates, as do the results from \texttt{JudgeIt}. Readers should note that, with only 18 districts, too much weight should not be placed on the exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures. Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay. We have devoted more space to partisan bias than to the other measures of gerrymandering reported in this essay due to it's complexity than to the other more straightforward measures.} 

@@

    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}). Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.

@@

    In our view, none of these four measures standing alone can be taken to be proof of gerrymandering. Rather we see them as potentially reinforcing indicators. Our own analyses have found hypotheticals where these measures can generate either a false positive or a false negative, but we have also found that consistent bias in the same direction is unlikely when we use multiple indicators (data omitted for space reasons). We will return to a discussion of the results later in this essay. Here we point out, however, all four indicators point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan -- namely that it is a blatant pro-Republican partisan gerrymander.
    \subsection*{Evaluation of the 2011 enacted map}

@@

        \subsubsection*{Partisan results in actual congressional elections held under the 2011 map}

@@

    \subsection*{Evaluating district plans using Statewide elections to provide a normal vote baseline}

@@

    In the next section we introduce a potentially important distinction between ``\textit{neutral plans}'' and ``\textit{fair plans}'', and we consider some important aspects of Pennsylvania's electoral geography.
